A programmer friendly communication framework

### 2024-03-10 2024_paper/paper/msglink_framework.tex
improved title
implemented changes from markus

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{caption}
% next 2 needed for code blocks
\usepackage{listings}
%\usepackage{etoolbox}  % needed for number detection, but didn't work reliably
% for URLs
\usepackage{url}

@@

\title{Programmer friendly communication framework}

@@

TCP stands for \textbf{Transmission Control Protocol} and is a connection based protocol, meaning there will be a standing connection between two hosts. Before any data can be sent over TCP, a connection has to be established during an initial handshake between both communication parties. This is done automatically in the background. Once established, a TCP connection provides a pipe-like environment for transmitting a continuous stream of data bytes. The data sent is guaranteed to be received in the same order (without missing bytes in between). However, since TCP "streams" bytes, it does guarantee that any block of bytes sent at once is received in one piece.

UDP stands for \textbf{User Datagram Protocol}. Unlike TCP, UDP does not require a connection to be established. UDP merely sends a block of data (user datagrams) all at once to a destination host within a single IP packet. There is no guarantee that the packet reaches the destination.

From here on, this paper will focus on reliable point-to-point communication between two hosts (although many of the concepts described later are also applicable to multicast communication) and on higher-level protocols and frameworks based on TCP.

@@

The work described in this paper aims to create a communication framework to solve problems for a rather specific set of requirements:

\begin{enumerate}
    \item Abstraction of low-level socket APIs, exposing the most important functionality with little code and automatically handle the rest in the background (according to common use-cases) including but not limited to automatic management of connection state, handling disconnects and reconnects
    \item Provide a way to exchange messages, which are blocks of data of a known (but still possibly dynamic) length, guaranteeing arrival in same grouping and order as sent
    \item Provide facilities to serialize and deserialize language-native data structures to easily send them over the network with the least code possible
    \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust received data to be in valid format.
    \item Provide additional, commonly used communication schemes and primitives such as metadata exchange or Remote Procedure Calls (RPCs). These further simplify common practices within network communication for the developer
\end{enumerate}

@@

The following section explains a typical protocol stack used for web apps, since one of the original primary use-cases of the work described in this paper is the communication between a single-page web-app and a high performance, low level server written in C++. This is used to implement (among other uses) the touchscreen UI to control FrenchBakery robots for the Botball competition. Since web browsers offer limited to no access to system socket APIs, a large part of the protocol stack is already defined.

@@

\textbf{TCP} as the underlying transportation layer. In an end-user application, this would typically only be used directly if a very high performance custom application protocol is key.

@@

\textbf{HTTP} builds on top of TCP. This protocol is used to transmit files and other chunks of data (such as JSON documents) between servers or between servers and a browser. One limitation is that HTTP is (mostly) unidirectional. HTTP works by the client connecting to a server using TCP and then sending an \textbf{HTTP Request}. The end of the request is identified by a CRLF sequence (an empty line of text). Depending on the request details, the server may then answers with some status code, response headers and finally the requested data. \cite{http-mdn}

@@

The WebSocket protocol however doesn't expose the raw stream socket to the programmer, instead it introduced the concept of messages, hereby implementing point 2) of the goals for a communication framework. A message is a self contained chunk of data (typically a string, raw binary is also possible) the transmission of which over the socket is transparent to the user. When the programmer sends messages, they are guaranteed to arrive at the other communication party in one piece and in order. \cite{b3}

@@

\section{Problems with current offer}

Although this protocol stack satisfies the most important goals listed, it still has some problems to be solved, namely the following:

\begin{enumerate}
    \item While the event system provided by Socket.IO is useful for many applications, sometimes responses to events are required. This is often times named a \textbf{Remote Procedure Call} and needs to be implemented manually for every event:
    \begin{itemize}
        \item define and keep track of a separate request and response event
        \item add an identifying element to request and response data to identify which response belongs to which request and prevent mixing up responses if multiple requests are sent out in short succession
        \item keep a list of open requests to handle responses once received.
    \end{itemize}
    \item While Socket.IO provides a very clean and feature-complete API for some programming languages (like Python and JavaScript), it doesn't do so for others like C++ which is only supported for clients. While there are libraries to create C++ WebSocket servers and clients (like websocket++ which is used in this paper), they require a lot of code (to some part due to the nature of the language) and tedious state management when used directly. Additionally, the event system would need to be implemented manually for every event every time, which is tedious. \cite{b4}
    \item The data serialization and deserialization of Socket.IO is fully dynamic, meaning that before an event is actually processed, the code cannot possibly know the structure of the event data. This presents multiple problems:
    \begin{itemize}
        \item Deserialization of event data to language-native data structures is impossible for statically typed languages like C++, which negates the benefits of static typing. It requires the use of dynamic containers like maps to represent objects with strings as attribute names. Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its's type and finally convert it to the representative language-native type, which is tedious.
        \item Even in dynamic languages like JavaScript where objects can have dynamic attributes and variables are dynamically typed it is still helpful to perform the same validation in order to provide adequate error handling and avoid unexpected behaviors. There are also additional systems (TypeScript for JS and type hints for Python) that provide the benefits of static typing (such as rich IntelliSense) even though the underlying language is dynamic.
    \end{itemize}
\end{enumerate}

@@

\begin{itemize}
    \item \textbf{Events} work in the same way as they do with Socket.IO, with the exception of having to define them and their data structure statically.
    \item \textbf{Remote Function Calls (RFC)} can be viewed as two events with opposite directions. One "call" event and one "result" event. RFCs work as the name implies. The called (receiving) side of the communication defines a function taking some input data (statically defined by schema) and returning some result data (also defined by a schema). Since a traditional function cannot yield multiple results, there can only be one such handler function. The other side of the communication can remotely call the function over the network, using asynchronous abilities of a language if available. This way, the remote function call can be used almost as seamlessly as a local one. The framework automatically handles the matching of results to pending calls so data is not mixed up.

@@

The following is a compressed code example of how one might define a msglink link in C++. The explanation of the code example is beyond the scope of this paper.

@@

A further limitation is the point-to-point nature of the framework. There always needs to be a msglink client connected to a msglink server, although a server can simultaneously have connections to multiple clients and can therefore distribute messages between them. At this time, it is not planned to directly implement support for broadcast or multicast communication into msglink, even on the server side (which Socket.IO does support), since this was not part of the original goal. For applications where many hosts need to communicate with each other directly in a mesh-like topology, there are other protocols better suited for the purpose that don't have the limitation of the browser. An example is DDS which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{caption}
% next 2 needed for code blocks
\usepackage{listings}
%\usepackage{etoolbox}  % needed for number detection, but didn't work reliably
% for URLs
\usepackage{url}
%used to to get the \hl{text} command for highlighting text - used by markus for feedback
\usepackage{soul}

@@

\title{A programmer friendly communication framework}

@@

TCP stands for \textbf{Transmission Control Protocol} and is a connection based protocol, meaning there will be a standing connection between two hosts. Before any data can be sent over TCP, a connection has to be established during an initial handshake between both communication parties. This is done automatically in the background. Once established, a TCP connection provides a pipe-like environment for transmitting a continuous stream of data bytes. The data sent is guaranteed to be received in the same order (without missing bytes in between). However, since TCP "streams" bytes, it does guarantee that any segment (group of bytes) sent with one system call is received in one piece.

UDP stands for \textbf{User Datagram Protocol}. Unlike TCP, UDP does not require a connection to be established. UDP merely sends a segment of data (user datagrams) to a destination host within a single IP packet. There is no guarantee that the packet reaches the destination.

From here on, this paper will focus on reliable point-to-point communication between two hosts and on higher-level protocols and frameworks based on TCP (although many of the concepts described later are also applicable to multicast communication).

@@

The work described in this paper aims to create a communication framework with the following set of goals:

\begin{enumerate}
    \item Abstraction of low-level socket APIs, exposing the most important functionality with little code and automatically handle the rest in the background (according to common use-cases) including but not limited to automatic management of connection state, handling disconnects and reconnects
    \item Provide a way to exchange messages, which are blocks of data of a known (but still possibly dynamic) length, guaranteeing arrival in same grouping and order as sent
    \item Provide facilities to serialize and deserialize language-native data structures to easily send them over the network with the least code possible
    \item Provide data validation for sent and received messages according to programmer defined schemas (in language native format as far as possible), so the user code can trust, that received data will be in the expected format.
    \item Provide additional, commonly used communication schemes and primitives such as metadata exchange or Remote Procedure Calls (RPCs). These further simplify common practices within network communication for the developer
\end{enumerate}

@@

The following section explains a typical protocol stack used for web apps, since one of the original primary use-cases of the work described in this paper is the communication between a single-page web-app and a high performance, low level server written in C++. This is used to implement (among other uses) the touchscreen UI to control FrenchBakery robots for the Botball competition. Since web browsers offer limited to no access to system socket APIs, a large part of the selected protocol stack is already defined by what browser APIs offer.

@@

\textbf{TCP} is selected as the underlying transportation layer. In an end-user application, this would typically only be used directly if a very high performance custom application protocol is key.

@@

\textbf{HTTP} builds on top of TCP. This protocol is used to transmit files and other chunks of data (such as JSON documents) between servers or between servers and a browser. One limitation is that HTTP is (mostly) unidirectional. HTTP works by the client connecting to a server using TCP and then sending an \textbf{HTTP Request}. The end of the request is identified by a CRLF sequence (an empty line of text). Depending on the request details, the server may then answer with some status code, response headers and finally the requested data. \cite{http-mdn}

@@

The WebSocket protocol however doesn't expose the raw stream socket to the programmer, instead it introduces the concept of messages, hereby implementing point 2) of the goals for a communication framework. A message is a self contained chunk of data (typically a string, raw binary is also possible) the transmission of which over the socket is transparent to the user. When the programmer sends messages, they are guaranteed to arrive at the other communication party in one piece and in order. \cite{b3}

@@

\section{Problems with current stack}

Although this protocol stack satisfies the most important goals listed, it still has the following problems:

\begin{enumerate}
    \item While the event system provided by Socket.IO is useful for many applications, sometimes responses to events are required. This is typically named a \textbf{Remote Procedure Call} and needs to be implemented manually for every event:
    \begin{itemize}
        \item define and keep track of a separate request and response event
        \item add an identifying element to request and response data to identify which response belongs to which request and prevent mixing up responses if multiple requests are sent out in short succession
        \item keep a list of open requests to handle responses once received.
    \end{itemize}
    \item While Socket.IO provides a very clean and feature-complete API for some programming languages (like Python and JavaScript), it doesn't do so for others like C++ which is only supported for clients. While there are libraries to create C++ WebSocket servers and clients (like websocket++ which is used in this paper), they require a lot of code (to some part due to the nature of the language) and tedious state management when used directly. Additionally, the event system would need to be implemented manually for every event every time, which is tedious. \cite{b4}
    \item The data serialization and deserialization of Socket.IO is fully dynamic, meaning that before an event is actually processed, the code cannot possibly know the structure of the event data. This presents multiple problems:
    \begin{itemize}
        \item Deserialization of event data to language-native data structures is impossible for statically typed languages like C++, which negates the benefits of static typing. It requires the use of dynamic containers like maps to represent objects with strings as attribute names. Since the containers need to be fully dynamic, the programmer first needs to write code to check if an expected attribute exists, assert its's type and finally convert it to the representative language-native type, which is tedious.
        \item Even in dynamic languages like JavaScript where objects can have dynamic attributes and variables are dynamically typed it is still helpful to perform the same validation in order to provide adequate error handling and avoid unexpected behaviors. There are also additional systems (TypeScript for JS and type hints for Python) that provide the benefits of static typing (such as rich IntelliSense) even though the underlying language is dynamic.
    \end{itemize}
\end{enumerate}

@@

\begin{itemize}
    \item \textbf{Events} work in the same way as they do with Socket.IO, with the exception of having to define them and their data structure statically.
    \item \textbf{Remote Function Calls (RFCs)} can be viewed as two events with opposite directions. One "call" event and one "result" event. RFCs work as the name implies. The called (receiving) side of the communication defines a function taking some input data (statically defined by schema) and returning some result data (also defined by a schema). Since a traditional function cannot yield multiple results, there can only be one such handler function. The other side of the communication can remotely call the function over the network, using asynchronous abilities of a language if available. This way, the remote function call can be used almost as seamlessly as a local one. The framework automatically handles the matching of results to pending calls so data is not mixed up.

@@

The following is a compressed code example of how one might define a msglink link in C++, which is significantly shorter than implementing a comparable solution manually using WebSocket++. The explanation of the code example is beyond the scope of this paper.

@@

A further limitation is the point-to-point nature of the framework. There always needs to be a msglink client connected to a msglink server, although a server can simultaneously have connections to multiple clients and can therefore distribute messages between them. At this time, it is not planned to directly implement support for broadcast or multicast communication into msglink, even on the server side (which Socket.IO does support), since this was not part of the original goal. For applications where many hosts need to communicate with each other directly in a mesh-like topology, there are other protocols better suited for the purpose that don't have the limitation of the browser. An example is the \textbf{Data Distribution Serivce (DDS)} which is used by ROS2 (Robot Operating System 2) and is therefore common in robotics.
